Implicit Neural Representation for Potential Field Geophysics

### 2023-10-13 discussion.tex
Final touch ups for typesetting and wording

\section{Image quality assessment for potential field grids}
Model performance is quantified using image quality assessments (IQA) such as peak signal-to-noise ratio (PSNR), structural similarity index (SSIM), feature similarity index (FSIM), or many others.
For supervised tasks, a full reference metric can be calculated between the super-resolution prediction and the high-resolution target.
In the works presented in \Cref{ch:paper1,ch:paper2,ch:paper3}, PSNR and SSIM are used to quantify the performance of the proposed methods in transforming low-resolution grids to high resolution.
However, properties of magnetic grid data are distinct to the images for which these metrics are designed.
%   While SSIM uses structural information present in the Fourier transformed input in combination with local gradient magnitude for value accuracy,
Consistently high SSIM scores are assigned to grids that are perceptually poor matches to the target.
High SSIM values are noted in other geophysical grid investigations which report SSIM metrics \parencite{wangDeeplearningbasedSeismicData2018,bavandsavadkoohiHighresolutionAeromagneticMap2023}.
An investigation to identify a more appropriate IQA for potential field grids is required, which accounts for the distinct nature of these data and better reflects the predicted quality of geological interpretation.
This may be benefit by automated analysis methods as mentioned above.

>>>>

\section{Image quality assessment for potential field grids}
Model performance is quantified using image quality assessments (IQA) such as peak signal-to-noise ratio (PSNR), structural similarity index (SSIM), feature similarity index (FSIM), or many others.
For supervised tasks, a full reference metric can be calculated between the super-resolution prediction and the high-resolution target.
In the works presented in \Cref{ch:paper1,ch:paper2,ch:paper3}, PSNR and SSIM or FSIM  are used to quantify the performance of the proposed methods in transforming low-resolution grids to high resolution.
However, properties of magnetic grid data are distinct to the images for which these metrics are designed.
%   While SSIM uses structural information present in the Fourier transformed input in combination with local gradient magnitude for value accuracy,
Consistently high SSIM and FSIM scores are assigned to grids that are perceptually poor matches to the target, including the low-resolution grids.
High SSIM values are noted in other geophysical grid investigations which report SSIM metrics \parencite{wangDeeplearningbasedSeismicData2018,bavandsavadkoohiHighresolutionAeromagneticMap2023}.
An investigation to identify a more appropriate IQA for potential field grids is required, which accounts for the distinct nature of these data and better reflects the predicted quality of geological interpretation.
This may be benefit by the automated analysis methods also mentioned above.

### 2023-10-13 paper2.tex
Final touch ups for typesetting and wording

Local texture estimator (LTE) \parencite{leeLocalTextureEstimator2022} is implemented in this work for its higher performance in the task of image super-resolution.
The encoder in LTE can use the feature extraction capability of traditional SR CNNs such as EDSR \parencite{limEnhancedDeepResidual2017} or RDN \parencite{zhangResidualDenseNetwork2018}, or a SwinIR \parencite{liangSwinIRImageRestoration2021} transformer-based encoder.
In the original implementation of the SR CNN networks, the extracted image features are upsampled and returned to the image domain by convolutional layers.
In LTE, the latent extracted features are further processed by the LTE network component before reconstruction to image space by an MLP\@.
An overview of the structure of the LTE network is shown in \Cref{fig:ltenet}.
Specifics of the LTE implementation to investigate the performance of SR for geophysical grids from a range of geological domains are detailed in \Cref{sec:2lte}.

@@

\subsection{Network selection}
\label{sec:2lte}
Local Texture Estimator \parencite{leeLocalTextureEstimator2022} is implemented to super-resolve synthetic and real potential field geophysical data with low resolution resulting from line spacing.
The LTE component of the LTE network transforms the encoded latent features to frequency and amplitude features using trainable convolutions.
Both the amplitude and frequency features are then upsampled using nearest neighbour filter interpolation.
The frequency components are combined with phase information, which is extracted from the input by an additional MLP, and the resulting features are passed through a sinusoidal activation function before being combined with the amplitude features.
Finally, the decoding MLP restores these frequency domain features to the spatial image domain.
To precondition the network toward learning high-resolution features, a long-range skip connection with bilinear upscaling passes low-resolution content common between the upsampled LR and the HR counterpart to the output of the MLP\@.
Because total magnetic intensity data are single channel, the input and output channel count are set to one.
The SwinIR feature encoder implementation by is used, and the MLP decoder retains the same configuration as the authors \parencite{leeLocalTextureEstimator2022}.

\section{Results}
\label{sec:2results}
Due to the dynamic range of magnetic anomaly grids and the relative importance of structural accuracy for gridded data, image quality assessment methods typically used for RGB images such as peak signal-to-noise ratio or mean-squared error are an inadequate quantitative measure of upscaling performance for geophysics.
In addition to mean-square error, FSIM \parencite{linzhangFSIMFeatureSimilarity2011} is used to quantify the quality of the super-resolved image.
FSIM is a full reference image similarity metric that quantifies structural similarity, rather than individual cell values, with highly similar structural content causing the metric to approach 1.0.
%TODO complain about SSIM Structural Similarity (SIM) \parencite{wangImageQualityAssessment2004}
Model performance is quantified using SSIM in \Cref{tab:2metrics}.
It is seen that fine-tune training increases FSIM score for the task of super-resolution for aeromagnetic grids, over the baseline model.

@@

    \caption[Accuracy Metrics]{Mean (\(\mu{}\)) and standard deviation (\(\sigma{}\)) of the FSIM metric for each upsampling method over the test set.}
    \label{tab:2metrics}
\end{table}

Qualitative results are shown to compare the outputs from the baseline model that is trained with synthetic data, and the fine-tuned model trained with open access survey data.

>>>>

Local texture estimator (LTE) \parencite{leeLocalTextureEstimator2022} is implemented in this work for its performance in the task of image super-resolution.
The encoder in LTE can use the feature extraction capability of traditional SR CNNs such as EDSR \parencite{limEnhancedDeepResidual2017} or RDN \parencite{zhangResidualDenseNetwork2018}, or a SwinIR \parencite{liangSwinIRImageRestoration2021} transformer-based encoder.
In the original implementation of the SR CNN networks, the extracted image features are upsampled and returned to the image domain by convolutional layers.
In LTE, the latent extracted features are further processed by the LTE network component before reconstruction to image space by an MLP\@.
An overview of the structure of the LTE network is shown in \Cref{fig:ltenet}.
Specifics of the LTE implementation to investigate the performance of SR for geophysical grids from a range of geological domains are detailed in \Cref{sec:2lte}.

@@

\subsection{Network selection}
\label{sec:2lte}
LTE \parencite{leeLocalTextureEstimator2022} is implemented to super-resolve synthetic and real potential field geophysical data with low resolution resulting from line spacing.
The LTE component of the LTE network transforms the encoded latent features to frequency and amplitude features using trainable convolutions.
Both the amplitude and frequency features are then upsampled using nearest neighbour filter interpolation.
The frequency components are combined with phase information, which is extracted from the input by an additional MLP, and the resulting features are passed through a sinusoidal activation function before being combined with the amplitude features.
Finally, the decoding MLP restores these frequency domain features to the spatial image domain.
To precondition the network for learning high-resolution features, a long-range skip connection with simple upscaling passes low-resolution content common between the upsampled LR and the HR target to the output of the MLP\@.
Because TMI data are single channel, the input and output channel count are set to one.
The SwinIR encoder implementation by is used, and the MLP decoder retains the same configuration as the authors \parencite{leeLocalTextureEstimator2022}.

\section{Results}
\label{sec:2results}
Due to the variable dynamic range of magnetic anomaly grids and the relative importance of structural accuracy for gridded data, image quality assessment methods typically used for RGB images such as peak signal-to-noise ratio or mean-squared error are an inadequate quantitative measure of upscaling performance for geophysics.
In addition FSIM \parencite{linzhangFSIMFeatureSimilarity2011} is used to quantify the quality of the super-resolved image.
FSIM is a full reference image similarity metric that quantifies structural similarity, rather than individual cell values, with highly similar structural content causing the metric to approach 1.0.
%TODO complain about SSIM Structural Similarity (SIM) \parencite{wangImageQualityAssessment2004}
Model performance is quantified using FSIM in \Cref{tab:2metrics}.
It is seen that fine-tune training increases FSIM score for the task of super-resolution for aeromagnetic grids, over the baseline model.

@@

    \caption[Accuracy Metrics]{Mean (\(\mu{}\)) and standard deviation (\(\sigma{}\)) of the FSIM metric for each upsampling method, for each test set.
    While the performance of the fine-tuned model is lower on synthetic data, the performance on real-world aeromagnetic grids is increased.}
    \label{tab:2metrics}
\end{table}

Qualitative results are shown to provide examples of structural improvement when using the fine-tuned model trained with open access survey data, compared to the baseline model trained with synthetic data.

### 2023-10-13 paper3.tex
Final touch ups for typesetting and wording

\begin{landscape}
    \begin{figure}[hbtp]
        \centering
        \includegraphics[width=1.0\linewidth]{fig/p3/P864_grid_comparison_40m.pdf}
        \caption[Grid prediction at the nominal altitude]{Comparison between the reference grid processed by Geoscience Australia (left) and the constant level grid queried from the INR at \qty{40}{\m} (middle). The residuals are shown for areas containing reference data (right), with a range of ±2 standard deviations of values in the reference grid.}
        \label{fig:grid40}
    \end{figure}
\end{landscape}

@@

\begin{figure}[hbtp]
    \centering{}
    \includegraphics[width=1\linewidth]{fig/p3/P864_dv_comparison.pdf}
    \caption[Vertical derivative]{Vertical derivative grids queried from AD differentiated INR (left) and calculated numerically (right) on the \qty{40}{\m} level grid queried from the INR\@.
        The proposed method has poor vertical gradient representation, possibly due to the single height sampled in aeromagnetic survey design.}
    \label{fig:vertgrad}
\end{figure}

@@

Inspired by the approach of \textcite{raissiPhysicsinformedNeuralNetworks2019}, a geophysics-informed network could include a Laplace criterion,
\begin{equation}
    \label{eqn:crilaplace}
    \mathcal{L}_{PDE} = \frac{1}{S}\sum_{i=1}^{S} \left(\frac{\partial{}^2f}{\partial{}x_{i}^2} + \frac{\partial{}^2f}{\partial{}y_{i}^2} + \frac{\partial{}^2f}{\partial{}z_{i}^2}\right)^2,
\end{equation}
where \(S_i\) indicates the \(i^{th}\) sample point of the field survey \(\phi(x, y, z)\), implicitly represented by the model \(f\).
This criterion penalises outputs where Laplace's equation is not satisfied because the sum of partial second derivatives is not close to zero.

>>>>

\begin{landscape}
    \begin{figure}[hbtp]
        \centering
        \includegraphics[width=1.0\linewidth]{fig/p3/P864_grid_comparison_40m.pdf}
        \caption[Grid prediction at the nominal altitude]{Comparison between the reference grid processed by Geoscience Australia (left) and the constant level grid queried from the INR at \qty{40}{\m} (middle). The residuals are shown (right) for areas containing reference data, with a range of ±2 standard deviations of values in the reference grid.}
        \label{fig:grid40}
    \end{figure}
\end{landscape}

@@

\begin{figure}[hbtp]
    \centering{}
    \includegraphics[width=1\linewidth]{fig/p3/P864_dv_comparison.pdf}
    \caption[Vertical derivative]{Vertical derivative grids calculated numerically on the \qty{40}{\m} level grid queried from the INR (left) and queried from AD differentiated INR (right).
        The proposed method has poor vertical gradient representation, possibly due to the low range of altitudes when sampling in aeromagnetic survey design.}
    \label{fig:vertgrad}
\end{figure}

@@

Inspired by the approach of \textcite{raissiPhysicsinformedNeuralNetworks2019}, a geophysics-informed network could include a Laplace criterion,
\begin{equation}
    \label{eqn:crilaplace}
    \mathcal{L}_{PDE} = \frac{1}{S}\sum_{i=1}^{S} \left(\frac{\partial{}^2f}{\partial{}x_{i}^2} + \frac{\partial{}^2f}{\partial{}y_{i}^2} + \frac{\partial{}^2f}{\partial{}z_{i}^2}\right)^2,
\end{equation}
where \(S\) indicates the number of sample points in the field survey \(\phi(x, y, z)\), implicitly represented by the model \(f\).
This criterion penalises outputs where Laplace's equation is not satisfied because the sum of partial second derivatives is not close to zero.
